Manuscript Template for a Data Analysis Project

### 2023-02-03 products/manuscript/MANU.qmd
commit

# Summary/Abstract
_Write a summary of your project._

@@

## General Background Information
_Provide enough background on your topic that others can understand the why and how of your analysis_

## Description of data and data source
The source of

## Questions/Hypotheses to be addressed
Question: Can HIV death rates be predicted by social determinants of health like poverty and/ or by HIV related outcomes (new diagnoses, prevalence PrEP use, etc?)

@@

_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._

## Data aquisition
_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._
Loading the datasets from OASIS for deaths and population (Hispanic, NH White, NH Black) for persons ages 13-59 from 2018 to 2020.
```{r warning=FALSE}
setwd("../../")
his_deaths <- read_xlsx("data/raw_data/hivdeaths_his_1820.xlsx", skip = 1)
nhw_deaths <- read_xlsx("data/raw_data/hivdeaths_nhw_1820.xlsx", skip = 1)
nhb_deaths <- read_xlsx("data/raw_data/hivdeaths_nhb_1820.xlsx", skip = 1)

@@

Loading dataset from AIDSVu for social determinants of health (2020) and new HIV diagnoses from 2018 to 2020.
```{r warning=FALSE}
setwd("../../")
sdoh <- read_xlsx("data/raw_data/AIDSVu_County_SDOH_2020.xlsx", skip = 3)
sdoh <- sdoh %>%  filter(State == "Georgia")

@@

Let's load data from Georgia Data on Poverty rates in Georgia by race/ ethnicity for the period 2016-2020.
```{r warning = false}
setwd("../../")
pov <- read_xlsx("data/raw_data/economics_22.xlsx", sheet = "poverty")

@@

We'll just be using the data from the "2016-2020 Persons Below Poverty Level" by race/ethnicity. We'll also change the text formatting of County to match the other data frames.
```{r}
pov <- pov %>%  select(1, starts_with("2016-2020 Persons Below Poverty Level"))

@@

Let's do some cleaning on the OASIS data. We'll remove extra columns and rows, as well as rows with "Selected ages/sexes total", pivot Year and Deaths into long format, and create the new variable for race/ethnicity.
```{r}
his_deaths <- his_deaths[c(-1,-2387:-2414),-7]
his_deaths <- his_deaths %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>%
  pivot_longer(c(4:6), names_to = "Year", values_to = "HIV_Deaths") %>% 
  mutate("Race_Ethnicity" = "Hispanic")

@@

Next, we'll merge the datasets into one dataframe, and remove the original datasets from the environment.
```{r}
deaths <- full_join(his_deaths, nhw_deaths)
deaths <- full_join(deaths, nhb_deaths)

@@

Next, let's clean up the dataset for population using a similar method.
```{r}
his_pop <- his_pop[c(-1,-2387:-2412),-7]
his_pop <- his_pop %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>% 
  pivot_longer(c(4:6), names_to = "Year", values_to = "Population") %>% 
  mutate("Race_Ethnicity" = "Hispanic")

@@

Merging datasets and cleaning the original from the environment.
```{r}
pop <- full_join(his_pop, nhw_pop)
pop <- full_join(pop, nhb_pop)

@@

Next, let's merge the two OASIS dataframes.
```{r}
df <- full_join(deaths, pop, by = c("Geography", "Sex", "Age", "Year", "Race_Ethnicity"))
```

Next, let's create a new variable for Public Health District, change some variable to factors, and rename the Age variable and Geography variable.
```{r}
df <- df %>% rename(County = Geography) %>% 
  mutate(District = case_when(County %in% c("Bartow", "Catoosa", "Chattooga", "Dade", "Floyd", "Gordon", "Haralson", "Paulding", "Polk", "Walker") ~ "1-1",
 County %in% c("Cherokee", "Fannin", "Gilmer", "Murray", "Pickens", "Whitfield") ~ "1-2",
 County %in% c("Banks", "Dawson", "Forsyth", "Franklin", "Habersham", "Hall", "Hart", "Lumpkin", "Rabun", "Stephens", "Towns", "Union", "White") ~ "2",
 County %in% c("Cobb", "Douglas") ~ "3-1",
 County %in% c("Fulton") ~ "3-2",
 County %in% c("Clayton") ~ "3-3",
 County %in% c("Gwinnett", "Newton", "Rockdale") ~ "3-4",
 County %in% c("DeKalb") ~ "3-5",
 County %in% c("Butts", "Carroll", "Coweta", "Fayette", "Heard", "Henry", "Lamar", "Meriwether", "Pike", "Spalding", "Troup", "Upson") ~ "4",
 County %in% c("Bleckley", "Dodge", "Johnson", "Laurens", "Montgomery", "Pulaski", "Telfair", "Treutlen", "Wheeler", "Wilcox") ~ "5-1",
 County %in% c("Baldwin", "Bibb", "Crawford", "Hancock", "Houston", "Jasper", "Jones", "Monroe", "Peach", "Putnam", "Twiggs", "Washington", "Wilkinson") ~ "5-2",
 County %in% c("Burke", "Columbia", "Emanuel", "Glascock", "Jefferson", "Jenkins", "Lincoln", "McDuffie", "Richmond", "Screven", "Taliaferro", "Warren", "Wilkes") ~ "6",
 County %in% c("Chattahoochee", "Clay", "Crisp", "Dooly", "Harris", "Macon", "Muscogee", "Marion", "Quitman", "Randolph", "Schley", "Stewart", "Sumter", "Talbot", "Taylor", "Webster") ~ "7",
 County %in% c("Ben Hill", "Berrien", "Brooks", "Cook", "Echols", "Irwin", "Lanier", "Lowndes", "Tift", "Turner") ~ "8-1",
 County %in% c("Baker", "Calhoun", "Colquitt", "Dougherty", "Decatur", "Early", "Grady", "Lee", "Miller", "Mitchell", "Seminole", "Terrell", "Thomas", "Worth") ~ "8-2",
 County %in% c("Bryan", "Camden", "Chatham", "Effingham", "Glynn", "Liberty", "Long", "McIntosh") ~ "9-1",
 County %in% c("Appling", "Atkinson", "Bacon", "Brantley", "Bulloch", "Candler", "Charlton", "Clinch", "Coffee", "Evans", "Jeff Davis", "Pierce", "Tattnall", "Toombs", "Ware", "Wayne") ~ "9-2",
 County %in% c("Barrow", "Clarke", "Elbert", "Greene", "Jackson", "Madison", "Morgan", "Oconee", "Oglethorpe", "Walton") ~ "10")) %>% 
   mutate(District = as.factor(District)) %>% 
   mutate(Sex = as.factor(Sex)) %>% 
   mutate(Age = as.factor(Age)) %>% 
   mutate(County = as.factor(County)) %>% 
   mutate(Race_Ethnicity = as.factor(Race_Ethnicity)) %>% 
   rename(Age_Group = Age) %>% 
   mutate(Age_Group = as.factor(Age_Group))
```

Let's clean up the Age_Group variable by removing text
```{r}
df$Age_Group <- str_replace(df$Age_Group, " Adolescence", "")
df$Age_Group <- str_replace(df$Age_Group, " Early Adulthood", "")
df$Age_Group <- str_replace(df$Age_Group, " Young Adulthood", "")
df$Age_Group <- str_replace(df$Age_Group, " Middle Adulthood", "")
```

```{r}
glimpse(df)
```
Things are looking good so far!

Now let's move onto the AIDSVu data. First, we'll merge all new HIV diagnoses from 2018-2020 into one dataframe
```{r}
dx <- full_join(dx18, dx19)
dx <- full_join(dx, dx20)
```

Let's remove columns for race/ethnicities other than Hispanic, NH White, and NH Black, as well as some other variables we won't be using in this analysis. We will also rename county and create the district variable. 
```{r}
dx <- dx %>% select(!contains(c("Asian", "American", "Multiracial", "Hawaiian", "MSM", "2013", "GEO ID", "State")))%>% 
  rename(County = "County Name")

@@

Things are looking pretty good so frame, so let's save these dataframes.
```{r}
saveRDS(df, "../../data/df.rds")
saveRDS(dx, "../../data/dx.rds")
saveRDS(sdoh, "../../data/sdoh.rds")
saveRDS(pov, "../../data/pov.rds")
```

Clear the environment and start fresh.
```{r}

```
## Statistical analysis
Explore central tendency across demographic groups. Fit a poisson regression model predicting death-to-case ratio from percentage with x SDOH variables and percentage of x demographic group.

@@

_Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project._

@@

_To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._

@@

_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here._

@@

## Summary and Interpretation
_Summarize what you did, what you found and what it means._

## Strengths and Limitations
_Discuss what you perceive as strengths and limitations of your analysis._

## Conclusions
_What are the main take-home messages?_

_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_

>>>>

# Summary/Abstract

*Write a summary of your project.*

@@

## General Background Information

*Provide enough background on your topic that others can understand the why and how of your analysis*

## Description of data and data source

The data used in this project was obtained from multiple publicly available sources. Data pertaining to counts of HIV-related deaths in for Hispanics, non-Hispanic Whites, and non-Hispanic Blacks and populations for these groups was obtained from [OASIS](https://oasis.state.ga.us/), the Georgia Department of Public Health's online data repository. This analyse will currently focus on the year 2018-2020, but may be expanded to 2016-2020 moving forward. The data on these years will be obtained and processed identically to ways later described. Date on new HIV diagnoses and SDOH, particularly related to HIV were obtained from [AIDSVu](https://aidsvu.org/resources/#/datasets). Data on poverty rates was obtained from [Georgia Data](https://georgiadata.org/), a data source provided by the the Carl Vinson Institute of Government at the University of Georgia. All data described was aggregated at the county level.

## Questions/Hypotheses to be addressed

Question: Can HIV death rates be predicted by social determinants of health like poverty and/ or by HIV related outcomes (new diagnoses, prevalence PrEP use, etc?)

@@

*Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement.*

## Data aquisition

*As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next.*

## Data import and cleaning

*Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along.* Loading the datasets from OASIS for deaths and population (Hispanic, NH White, NH Black) for persons ages 13-59 from 2018 to 2020.

```{r warning=FALSE}
setwd("../../")
his_deaths <- read_xlsx("data/raw_data/hivdeaths_his_1820.xlsx", skip = 1)
nhw_deaths <- read_xlsx("data/raw_data/hivdeaths_nhw_1820.xlsx", skip = 1)
nhb_deaths <- read_xlsx("data/raw_data/hivdeaths_nhb_1820.xlsx", skip = 1)

@@

Loading dataset from AIDSVu for social determinants of health (2020) and new HIV diagnoses from 2018 to 2020.

```{r warning=FALSE}
setwd("../../")
sdoh <- read_xlsx("data/raw_data/AIDSVu_County_SDOH_2020.xlsx", skip = 3)
sdoh <- sdoh %>%  filter(State == "Georgia")

@@

Let's load data from Georgia Data on Poverty rates in Georgia by race/ ethnicity for the period 2016-2020.

```{r warning = false}
setwd("../../")
pov <- read_xlsx("data/raw_data/economics_22.xlsx", sheet = "poverty")

@@

We'll just be using the data from the "2016-2020 Persons Below Poverty Level" by race/ethnicity. We'll also change the text formatting of County to match the other data frames.

```{r}
pov <- pov %>%  select(1, starts_with("2016-2020 Persons Below Poverty Level"))

@@

Let's do some cleaning on the OASIS data. We'll remove extra columns and rows, as well as rows with "Selected ages/sexes total", pivot Year and Deaths into long format, and create the new variable for race/ethnicity.

```{r}
his_deaths <- his_deaths[c(-1,-2387:-2414),-7]
his_deaths <- his_deaths %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>%
  pivot_longer(c(4:6), names_to = "Year", values_to = "HIV_Deaths") %>% 
  mutate("Race_Ethnicity" = "Hispanic")

@@

Next, we'll merge the datasets into one dataframe, and remove the original datasets from the environment.

```{r}
deaths <- full_join(his_deaths, nhw_deaths)
deaths <- full_join(deaths, nhb_deaths)

@@

Next, let's clean up the dataset for population using a similar method.

```{r}
his_pop <- his_pop[c(-1,-2387:-2412),-7]
his_pop <- his_pop %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>% 
  pivot_longer(c(4:6), names_to = "Year", values_to = "Population") %>% 
  mutate("Race_Ethnicity" = "Hispanic")

@@

Merging datasets and cleaning the original from the environment.

```{r}
pop <- full_join(his_pop, nhw_pop)
pop <- full_join(pop, nhb_pop)

@@

Next, let's merge the two OASIS dataframes.

```{r}
df <- full_join(deaths, pop, by = c("Geography", "Sex", "Age", "Year", "Race_Ethnicity"))
```

Next, let's create a new variable for Public Health District, change some variable to factors, and rename the Age variable and Geography variable.

```{r}
df <- df %>% rename(County = Geography) %>% 
  mutate(District = case_when(County %in% c("Bartow", "Catoosa", "Chattooga", "Dade", "Floyd", "Gordon", "Haralson", "Paulding", "Polk", "Walker") ~ "1-1",
 County %in% c("Cherokee", "Fannin", "Gilmer", "Murray", "Pickens", "Whitfield") ~ "1-2",
 County %in% c("Banks", "Dawson", "Forsyth", "Franklin", "Habersham", "Hall", "Hart", "Lumpkin", "Rabun", "Stephens", "Towns", "Union", "White") ~ "2",
 County %in% c("Cobb", "Douglas") ~ "3-1",
 County %in% c("Fulton") ~ "3-2",
 County %in% c("Clayton") ~ "3-3",
 County %in% c("Gwinnett", "Newton", "Rockdale") ~ "3-4",
 County %in% c("DeKalb") ~ "3-5",
 County %in% c("Butts", "Carroll", "Coweta", "Fayette", "Heard", "Henry", "Lamar", "Meriwether", "Pike", "Spalding", "Troup", "Upson") ~ "4",
 County %in% c("Bleckley", "Dodge", "Johnson", "Laurens", "Montgomery", "Pulaski", "Telfair", "Treutlen", "Wheeler", "Wilcox") ~ "5-1",
 County %in% c("Baldwin", "Bibb", "Crawford", "Hancock", "Houston", "Jasper", "Jones", "Monroe", "Peach", "Putnam", "Twiggs", "Washington", "Wilkinson") ~ "5-2",
 County %in% c("Burke", "Columbia", "Emanuel", "Glascock", "Jefferson", "Jenkins", "Lincoln", "McDuffie", "Richmond", "Screven", "Taliaferro", "Warren", "Wilkes") ~ "6",
 County %in% c("Chattahoochee", "Clay", "Crisp", "Dooly", "Harris", "Macon", "Muscogee", "Marion", "Quitman", "Randolph", "Schley", "Stewart", "Sumter", "Talbot", "Taylor", "Webster") ~ "7",
 County %in% c("Ben Hill", "Berrien", "Brooks", "Cook", "Echols", "Irwin", "Lanier", "Lowndes", "Tift", "Turner") ~ "8-1",
 County %in% c("Baker", "Calhoun", "Colquitt", "Dougherty", "Decatur", "Early", "Grady", "Lee", "Miller", "Mitchell", "Seminole", "Terrell", "Thomas", "Worth") ~ "8-2",
 County %in% c("Bryan", "Camden", "Chatham", "Effingham", "Glynn", "Liberty", "Long", "McIntosh") ~ "9-1",
 County %in% c("Appling", "Atkinson", "Bacon", "Brantley", "Bulloch", "Candler", "Charlton", "Clinch", "Coffee", "Evans", "Jeff Davis", "Pierce", "Tattnall", "Toombs", "Ware", "Wayne") ~ "9-2",
 County %in% c("Barrow", "Clarke", "Elbert", "Greene", "Jackson", "Madison", "Morgan", "Oconee", "Oglethorpe", "Walton") ~ "10")) %>% 
   mutate(District = as.factor(District)) %>% 
   mutate(Sex = as.factor(Sex)) %>% 
   mutate(Age = as.factor(Age)) %>% 
   mutate(County = as.factor(County)) %>% 
   mutate(Race_Ethnicity = as.factor(Race_Ethnicity)) %>% 
   rename(Age_Group = Age) %>% 
   mutate(Age_Group = as.factor(Age_Group))
```

Let's clean up the Age_Group variable by removing text

```{r}
df$Age_Group <- str_replace(df$Age_Group, " Adolescence", "")
df$Age_Group <- str_replace(df$Age_Group, " Early Adulthood", "")
df$Age_Group <- str_replace(df$Age_Group, " Young Adulthood", "")
df$Age_Group <- str_replace(df$Age_Group, " Middle Adulthood", "")
```

```{r}
glimpse(df)
```

Things are looking good so far!

Now let's move onto the AIDSVu data. First, we'll merge all new HIV diagnoses from 2018-2020 into one dataframe

```{r}
dx <- full_join(dx18, dx19)
dx <- full_join(dx, dx20)
```

Let's remove columns for race/ethnicities other than Hispanic, NH White, and NH Black, as well as some other variables we won't be using in this analysis. We will also rename county and create the district variable.

```{r}
dx <- dx %>% select(!contains(c("Asian", "American", "Multiracial", "Hawaiian", "MSM", "2013", "GEO ID", "State")))%>% 
  rename(County = "County Name")

@@

Things are looking pretty good so frame, so let's save these dataframes.

```{r}
saveRDS(df, "../../data/df.rds")
saveRDS(dx, "../../data/dx.rds")
saveRDS(sdoh, "../../data/sdoh.rds")
saveRDS(pov, "../../data/pov.rds")
```

Clear the environment and start fresh.

```{r}

```

## Statistical analysis

Explore central tendency across demographic groups. Fit a poisson regression model predicting death-to-case ratio from percentage with x SDOH variables and percentage of x demographic group.

@@

*Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project.*

@@

*To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*

@@

*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here.*

@@

## Summary and Interpretation

*Summarize what you did, what you found and what it means.*

## Strengths and Limitations

*Discuss what you perceive as strengths and limitations of your analysis.*

## Conclusions

*What are the main take-home messages?*

*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*
